---
title: "NHANES BMI Analysis Dashboard"
format: 
  html:
    theme: cosmo
    toc: true
runtime: shiny
---

## Introduction

This includes analyses of different sociodemographic factors and their impact on BMI for adult females (age 20-79) in the United States included in the 2017-2020 NHANES study.

## Analysis 1 - Comparison of BMI & Waist Circumference by Sex

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(NHANES)
library(haven)
library(stats)
library(broom)
library(tidyverse)

theme_set(theme_light())
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
url_2017_2020 <- "https://wwwn.cdc.gov/Nchs/Data/Nhanes/Public/2017/DataFiles/P_DEMO.xpt"

download.file(url_2017_2020, destfile = "DEMO_J.XPT", mode = "wb")

demo_2017_2020 <- read_xpt("DEMO_J.XPT")

if (!dir.exists("data/demographics")) {
  dir.create("data/demographics", recursive = TRUE)
}

saveRDS(demo_2017_2020, "data/demographics/demo_2017_2020.Rds")

```

```{r}
bmi_raw <- read_xpt("C:/Project B/Part2/ProjectBPart2/P_BMX.xpt")

if (!dir.exists("data")) {
  dir.create("data")
}
saveRDS(bmi_raw, "data/P_BMX.Rds")

demo_data <- as_tibble(demo_2017_2020)

saveRDS(demo_data, "data/demographics/demo_2017_2020.Rds")

bmi_data <- as_tibble(bmi_raw)

saveRDS(bmi_data, "data/P_BMX.Rds")

demo_data <- readRDS("data/demographics/demo_2017_2020.Rds")
bmi_data <- readRDS("data/P_BMX.Rds")

demo_bmi <- left_join(demo_data, bmi_data, by = "SEQN")
```

```{r}
analysisAA_data <- demo_bmi %>%
  filter(RIDAGEYR >= 18, RIDAGEYR <= 79,
         !is.na(BMXBMI), !is.na(BMXWAIST))
```

```{r plot-bmi-waist-circumference, echo=FALSE, fig.cap="Differences in BMI and Waist Circumference by Sex"}

# Data transformation
analysisAA_data <- analysisAA_data %>%
  mutate(sex = ifelse(RIAGENDR == 1, "Male", "Female")) %>%
  select(SEQN, RIAGENDR, BMXBMI, BMXWAIST, sex) %>%
  pivot_longer(cols = c(BMXBMI, BMXWAIST), names_to = "variable", values_to = "value") %>%
  mutate(variable = recode(variable, BMXBMI = "BMI", BMXWAIST = "Waist Circumference"))

# Plotting
p <- ggplot(analysisAA_data, aes(x = sex, y = value, fill = sex)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y") +  # Allows different y-scales for BMI and Waist Circumference
  labs(title = "Comparison of BMI and Waist Circumference by Sex",
       x = "Sex",
       y = "Measurement") +
  theme_minimal()

p 
```

## Analysis 2 - Comparison of BMI by Type of Insurance

```{r bmi-ridgeline-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Ridgeline Plot of BMI Distribution by Insurance Type"}

hi_raw <- read_xpt("C:/Project B/Part2/ProjectBPart2/P_HIQ.xpt")

if (!dir.exists("data")) {
  dir.create("data")
}
saveRDS(hi_raw, "data/P_HIQ.Rds")

hi_data <- as_tibble(hi_raw)

saveRDS(hi_data, "data/P_HIQ.Rds")

demo_bmi_hi <- left_join(demo_data, bmi_data, by = "SEQN") %>%
               left_join(hi_data, by = "SEQN")

filtered_dataC <- demo_bmi_hi %>%
  filter(RIAGENDR == 2,            
         RIDAGEYR >= 20,           
         RIDAGEYR <= 79) %>%       
  mutate(InsuranceType = case_when(
    HIQ032A == 1 ~ "Private",
    HIQ032B == 2 ~ "Medicare",
    HIQ032D == 4 ~ "Medicaid",
    TRUE ~ "Other"                 
  )) %>%
  filter(InsuranceType %in% c("Private", "Medicare", "Medicaid"))  

filtered_dataC <- filtered_dataC %>%
  filter(!is.infinite(BMXBMI), !is.na(BMXBMI))

tryCatch({
    insurance_bmi_ridgeline_plot <- ggplot(filtered_dataC, aes(x = BMXBMI, y = InsuranceType, fill = InsuranceType)) +
      ggridges::geom_density_ridges() +
      labs(title = "Ridgeline Plot of BMI Distribution by Insurance Type",
           x = "BMI", y = "Insurance Type") +
      theme_minimal() +
      scale_fill_brewer(palette = "Set2")
}, warning = function(w) {})

# Display the plot
insurance_bmi_ridgeline_plot
```

## Analysis 3 - Comparison of BMI by Healthcare Access

```{r bmi-bar-chart, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="BMI Data by Healthcare Access"}
access_raw <- read_xpt("C:/Project B/Part2/ProjectBPart2/P_HUQ.xpt")

if (!dir.exists("data")) {
  dir.create("data")
}
saveRDS(access_raw, "data/P_HUQ.Rds")

access_data <- as_tibble(access_raw)

saveRDS(access_data, "data/P_HUQ.Rds")

demo_bmi_access <- left_join(demo_data, bmi_data, by = "SEQN") %>%
               left_join(access_data, by = "SEQN")

filtered_dataD <- demo_bmi_access %>%
  filter(RIAGENDR == 2,                  
         RIDAGEYR >= 18, RIDAGEYR <= 79, 
         !is.na(HUQ030),                 
         HUQ030 %in% c(1, 2)) %>%        
  mutate(
    HealthcareAccess = recode(HUQ030, '1' = "Yes", '2' = "No"),
    Overweight = ifelse(BMXBMI >= 25, "Overweight", "Not Overweight")
  )

ggplot(filtered_dataD, aes(x = HealthcareAccess, fill = Overweight)) +
  geom_bar() +
  labs(title = "Count of Overweight Status by Healthcare Access",
       x = "Healthcare Access",
       y = "Count") +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal()
```

## Analysis 4 - Prediction of BMI by multiple factors

```{r coef-plot, echo=FALSE, fig.cap="Impact of various predictors on Log-Transformed BMI"}

analysis2 <- demo_bmi_hi %>%
  filter(RIDAGEYR >= 20, RIDAGEYR <= 79, RIAGENDR == "2")

analysis2 <- analysis2 %>%
  mutate(Education_Level = case_when(
    DMDEDUC2 %in% c(1, 2) ~ "Did not graduate HS",
    DMDEDUC2 == 3 ~ "HS Graduate",
    DMDEDUC2 %in% c(4, 5) ~ "College Graduate"
  ))

analysis2 <- analysis2 %>%
 mutate(InsuranceType = case_when(
    HIQ032A == 1 ~ "Private",
    HIQ032B == 2 ~ "Medicare",
    HIQ032D == 4 ~ "Medicaid",
    TRUE ~ "Other"                 
  )) %>%
  filter(InsuranceType %in% c("Private", "Medicare", "Medicaid"))

analysis2 <- analysis2 %>%
  mutate(Marital_Status = case_when(
    DMDMARTZ %in% c(1) ~ "Married",
    DMDMARTZ %in% c(2, 3) ~ "Not Married"
  ))

analysis2_final <- analysis2 %>%
  select(SEQN, RIDAGEYR, BMXBMI, InsuranceType, Education_Level, Marital_Status)

analysis2_cleaned <- analysis2_final %>%
  filter(!is.na(BMXBMI)) 

analysis2_cleaned <- analysis2_cleaned %>%
  filter(!is.na(Education_Level) & !is.na(Marital_Status))

analysis2_cleaned <- analysis2_cleaned %>%
  mutate(Log_BMI = log(BMXBMI))

set.seed(123)

train_index <- sample(1:nrow(analysis2_cleaned), size = 0.75 * nrow(analysis2_cleaned))

train_data <- analysis2_cleaned[train_index, ]
test_data <- analysis2_cleaned[-train_index, ]

model_fit <- lm(Log_BMI ~ InsuranceType + Education_Level + Marital_Status + RIDAGEYR, data = train_data)

coefs <- tidy(model_fit)

coefs <- coefs %>%
  mutate(Significant = ifelse(p.value < 0.05, "Significant", "Not Significant"))

ggplot(coefs, aes(x = term, y = estimate, fill = Significant)) + 
  geom_col(show.legend = TRUE) +
  coord_flip() +
  labs(title = "Impact of Predictors on Log-Transformed BMI",
       x = "Predictors",
       y = "Coefficient Estimate") +
  theme_minimal() +
  scale_fill_manual(values = c("Significant" = "steelblue", "Not Significant" = "gray"))

coefs$Significant <- factor(coefs$Significant, levels = c("Not Significant", "Significant"))
```
